---
title: 'MLR: Confounding and Interaction'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

## Interaction Example

Is there a relationship between age and birthweight?

```{r}
# read data
birth = read_csv("./lowbwt_ALL.csv") %>%
  mutate(smoke = factor(smoke), race = factor(race))

# birthweight vs age
ggplot(birth) + 
  geom_point(aes(x = age, y = bwt)) +
  theme_bw()

# SLR
slr = lm(bwt ~ age, data = birth)
summary(slr)

```

Interaction: the relationship between 2 variables is different (or varies) based on the level of a third variable

Does the relationship between age and birth weight vary by smoking status? 

```{r}
# birthweight vs age by smoking status
ggplot(birth, aes(x = age, y = bwt, color = smoke)) + 
  geom_point() +
  geom_smooth(method="lm", se=F, aes(group = smoke, color = smoke)) +
  theme_bw()

# fit model with interaction
interact = lm(bwt ~ age*smoke, data = birth)
summary(interact)

```

Stratify the anaysis by smoking status. 

* Smokers have a negative, but not statistically significant association b/w age and baby birth weight

* Non-Smokers have a positive, statistically significant association b/w age and baby birth weight


```{r}
# subset data
smokers = filter(birth, smoke == "1")
nonsmokers = filter(birth, smoke == "0")

# fit models
summary(lm(bwt ~ age, data = smokers))
summary(lm(bwt ~ age, data = nonsmokers))

```


## Confounding Example

To be a confounder the variable must be:

1) Associated with the outcome (bwt)

2) Associated with the exposure (age)

3) Not in the causal pathway between the exposure and outcome

Does race confound the relationship between age and birthweight?

Answer: The age coefficient reduced from 12.36 in SLR to 6.19 in MLR after adjusting for race (~50% reduction). 
We can conclude that race confounds the relationship between age and birth weight.


```{r}
# condition (1)
bwt_race = lm(bwt ~ race, data = birth)
summary(bwt_race)

# condition (2)
age_race = lm(age ~ race, data = birth)
summary(age_race)

# fit full model 
full = lm(bwt ~ age + race, data = birth)
summary(full)
anova(full)

```

Is there an interaction between age and race?

```{r}
# birthweight vs age by race
ggplot(birth, aes(x = age, y = bwt, color = race)) + 
  geom_point() +
  geom_smooth(method="lm", se=F, aes(group = race, color = race)) +
  theme_bw()

# check for interaction
summary(lm(bwt ~ age*race, data = birth))

# summarize across levels of race
anova(lm(bwt ~ age*race, data = birth))

```





